Local fitness and epistatic effects lead to distinct patterns of linkage disequilibrium in protein-coding genes

### 2022-06-03 manuscript/paper.tex
release 0.2
updates for minor revisions

\usepackage{lineno}
\linenumbers

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by a lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations. Using
whole-genome sequencing data from humans, I explore genome-wide patterns of LD
within protein-coding genes. I show that positive LD between missense mutations
within genes is driven by strong positive allele-frequency correlations between
pairs of mutations that fall within the same annotated conserved domain,
pointing to compensatory mutations or antagonistic epistasis as the prevailing
mode of interaction within conserved genic elements. LD between missense
mutations is reduced outside of conserved domains, as would expected under
Hill-Robertson interference. This variation in both mutational fitness effects
and selective interactions within protein-coding genes calls for more refined
inferences of the joint distribution of fitness and interactive effects, and
the methods presented here should prove useful in that pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions \citep[e.g.,][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
This distortion away from neutral expectations has been empirically documented
using sequencing data from an ever-growing set of study systems
\citep{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions
remain about the primary mode of interactions between multiple linked variants
and their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from single-locus expectations.
Under a multiplicative fitness model, where the fitness reduction of a genome
carrying multiple deleterious variants is equal to the product of the fitness
reduction of each independent mutation, we expect to see net linkage
disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
But for linked loci, those mutations are expected to segregate on different
haplotypes more often than together, leading to negative, or repulsion, LD,
although the extent of LD depends non-trivially on the strength of selection
and the probability of recombination separating loci
\citep{Hill1966-gv,McVean2000-ox}.

@@

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations in protein-coding genes and annotated conserved
domains. I show that while synonymous and missense variants display similar
slightly positive average LD within genes, for missense mutations this signal
is driven by correlations between pairs of mutations within, but not between or
outside of, protein-coding domains. This suggests an importance for
antagonistic epistasis or a prevalence of compensatory nonsynonymous mutations
within conserved elements.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
Puchta2016-rx, Steinberg2016-is}. For example, using the model system of the
TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
found evidence for synergistic epistasis, in which the combined effect of
multiple deleterious mutations on individual fitness was larger than would be
expected from multiplying the independent observed effects of each individual
mutation. The scale of mutation scanning experiments continues to increase,
promising greater resolution of the fitness landscape in such model systems
that can be compared to evolutionary theory \citep{Otwinowski2018-zk}.

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theoretical
predictions that linked negatively selected mutations will display negative LD
due to interference \citep{Hill1966-gv} and that epistasis will drive expected
LD away from zero, a number of recent studies have used patterns of LD within
classes of putatively selected variants to infer modes of selective
interactions.  \citet{Callahan2011-ac} observed that pairs of tightly linked
nonsynonymous mutations cluster together more than expected along the same
lineages in the \emph{Drosophilid} species complex, and that those clustered
mutations tend to preserve the charge of the protein and were in positive LD
compared to pairs of synonymous mutations at the same distance. From this, they
proposed that compensatory nonsynonymous variants are regularly tolerated and
maintained.  More recently, \citet{Taverner2020-lk} replicated this finding
across a diverse set of genera, showing that such epistatic interactions are
important for protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. This was interpreted as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutation
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their low allele frequencies mean that measurements of LD and other diversity
statistics are very noisy. Second, comparisons are based on theory with
limiting assumptions, such as steady-state demography, simple selection and
interaction models, or unlinked loci. To generate predictions under more
complex models, we rely on expensive forward simulations. Such simulations can
help build intuition and be used to test inference methods, but they do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g., conditioned on observing
        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution of
two-locus haplotypes (\(\Psi_n\), Figure~\ref{fig:TwoLocusFS}) under a general
model of selective interactions. This sampling distribution stores the expected
density or observed counts of pairs of biallelic loci with each possible
haplotype frequency configuration in a sample of size \(n\). Below, we compute
expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linked selected mutations, such as background
selection due to many linked variants, and individual-based forward simulations
are still needed for such scenarios (e.g.,
Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of
the distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution \(\Psi_n\). Here I focus on low-order LD
statistics including \(\E[D]\) and \(\E[D^2]\) and their decay with
recombination distance, as these are statistics that are commonly used to test
for interactions between loci. For pairs of biallelic loci, with alleles
labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
\(D=f_{AB}-f_A f_B\) is the standard covariance measure of LD, where \(f_{AB}\)
is the frequency of haplotypes carrying both \(A\) and \(B\), and \(f_A\) and
\(f_B\) are the marginal allele frequencies of \(A\) and \(B\).

Instead of unnormalized \(\E[D^2]\) and \(\E[D]\) I consider expectations for
\(\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\).  Such normalized statistics have two
benefits: first, the mutation rate cancels so that expectations are robust to
assumptions about the per-base mutation rate, and second, we can compare to
analytic expectations for these quantities under neutrality and constant
population size \citep{Ohta1971-yd}.

Below, I first consider the case of additive selection, Hill-Robertson
interference, and epistasis. I then explore the effect of dominance acting
within loci but without epistatis, and then describe a general diploid
selection model and consider gene-based dominance effects. In what follows I
focus on parameters where the strengths of selection and dominance at each
locus are equal, but note that the methods presented here allow for arbitrary
and unequal selection and dominance at the two loci. I also focus primarily on
weak to moderate negative selection (\(|2Ns| \approx 1 - 20\)), since this
range of selection leads to the strongest signals of interference
(Figure~\ref{fig:HillRobertson}) and is the parameter regime for which the
numerical appraoch is most accurate.

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations can cause negative
        signed LD in the absence of dominance and epistatic effects.
        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
        decreases below neutral expectations
        as the strength of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\).
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for somewhat stronger selection coefficients.
        Dashed lines show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]\) with recombination
        distance, and right panels (B and D) show expectations for the
        decay of \(\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]\). Dashed lines
        show neutral expectations.
        For both weak (\(s=-1/2N\), A and B) and moderate
        (\(s=-10/2N\), C and D) selection,
        antagonistic epistasis (\(\epsilon < 0\))
        gives rise to positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
    }
    \label{fig:epistasis}
\end{figure}

For mutations under additive selection (\(h=1/2\)) and no epistasis, we
recover the well known \citet{Hill1966-gv} interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients \(\gamma = O(1)\), or \(s \approx 1/2N_e\) (Figure
\ref{fig:HillRobertson}).
For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other \citep{McVean2000-ox}.

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \ref{fig:epistasis}).
Synergistic epistasis (in which the effect of two mutations together is larger than
the product of each individual mutation's effect) results in negative LD while
antagonistic epistasis (in which the combined effect is less than the product of
independent effects) results in positive LD, and large nonzero LD can occur
even when epistasis is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \ref{fig:epistasis}F even moderately deleterious mutations with
population-size-scaled selection coefficients of \(\gamma=-10\) show large mean
LD that extends to values of \(\rho\) much greater than 1 (for humans, assuming
roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
More strongly deleterious interacting mutations are expected to show large
signed LD over much larger recombination distances.

@@

\begin{figure}[tb!]
    \internallinenumbers
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C) Interference effects are most pronounced for recessive deleterious
        variants.
        (D, E) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        Selection parameters of \(sh=-0.1\) imply extremely strong selection
        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
        diploids at a single locus). The numerical appraoch for \(\Psi_n\)
        cannot handle such strong selection.
    }
    \label{fig:dominance}
\end{figure}

The effect of non-additive selection on correlations between mutations has
received increased attention recently. For example, \citet{Garcia2021-zn} used
large-scale forward simulations to explore how dominance impacts patterns of
LD, showing that LD depends on the magnitudes of both the selection and
dominance coefficients in a nonlinear way. \citet{Roze2021-cf} found an
analytic expression for LD between pairs of strongly deleterious mutations
under steady-state demography, showing that LD can be either positive or
negative depending on the strength of dominance.

The combined effect of the strength of selection and dominance on interference
is indeed nontrivial (Figure~\ref{fig:dominance}A), as observed by
\citet{Garcia2021-zn}. Some parameter regimes can cause strong negative LD
between pairs of negatively selected variants, with moderately selected
recessive variants having stronger signals of interference than additive
selection (Figure~\ref{fig:dominance}B). Unlike epistatic interactions, signed
LD decays rapidly with increasing distance between loci and is roughly zero for
\(\rho \gg 1\). For weakly selected mutations with \(\gamma=-1\), there is no
monotonic effect of the level of dominance on negative LD, with both recessive
and dominant pairs of mutations having more negative LD than the additive case.

Discrete simulations and the moments approach confirm the result from
\citet{Roze2021-cf}, that positive LD can occur for strong negative selection
and small values of \(h\) (Figure~\ref{fig:dominance}D, E). However, while the
analytic formula in \citet{Roze2021-cf} predicts that LD should be positive for
\(h<0.25\) and negative for \(h>0.25\), this appears to only hold in the limit
of strong selection (compare to Figure 1A in \citet{Roze2021-cf}). For moderate
to moderately strong selection, this threshold of \(h\) can be less than
\(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
mutations with interference strongest between pairs of partially recessive
mutations (Figure~\ref{fig:dominance}C).

\subsubsection{General selection and gene-based dominance}
\label{sec:general-selection}

>>>>

\usepackage{lineno}
%\linenumbers

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. Recent studies have used patterns of linkage
disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by a lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations. Using
whole-genome sequencing data from humans, I explore genome-wide patterns of LD
within protein-coding genes. I show that positive LD between missense mutations
within genes is driven by strong positive allele-frequency correlations between
mutations that fall within the same annotated conserved domain, pointing to
compensatory mutations or antagonistic epistasis as the prevailing mode of
interaction within conserved genic elements. LD between missense mutations is
reduced outside of conserved domains, as expected under Hill-Robertson
interference. This variation in both mutational fitness effects and selective
interactions within protein-coding genes calls for more refined inferences of
the joint distribution of fitness and interactive effects, and the methods
presented here should prove useful in that pursuit.

@@

Most new mutations that affect fitness are deleterious and will be eliminated
from a population. The amount of time that a deleterious mutation segregates
depends on the strength of selection against genomes that carry it, with very
damaging mutations kept at low frequencies and purged relatively rapidly. In
the time between mutation and fixation or loss, selected variants can
dramatically impact patterns of variation in nearby regions
\citep{Smith1974-em,Charlesworth1995-dq,Kim2000-on}. This distortion
away from neutral expectations has been empirically documented using sequencing
data from an ever-growing set of study systems
\citep[e.g.,][]{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions remain
about the primary mode of interactions between selected variants and their
joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from single-locus expectations.
Under a multiplicative fitness model, where the fitness reduction of a genome
carrying multiple deleterious variants is equal to the product of the fitness
reduction of each independent mutation, net linkage disequilibrium (LD) is
expected to be zero for unlinked sites \citep{Kondrashov1995-va}.  However,
mutations at linked loci are expected to segregate on different haplotypes more
often than together, leading to negative, or repulsion, LD, and the magnitude
of LD depends non-trivially on the strength of selection and the probability of
recombination separating loci \citep{Hill1966-gv,McVean2000-ox}.

@@

Here I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations in protein-coding genes and annotated conserved
domains. I show that while synonymous and missense variants display similar
slightly positive average LD within genes, for missense mutations this signal
is driven by correlations between pairs of mutations within, but not between or
outside of, conserved annotated elements. This suggests an importance for
antagonistic epistasis or a prevalence of compensatory nonsynonymous mutations
within conserved gene domains.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or protein
function is experimentally measured \citep{Romero2009-yi, Bank2015-vq,
Puchta2016-rx, Steinberg2016-is}. Using the model system of the
TEM-1 \(\beta\)-lactamase gene in \emph{E. coli}, \citet{Bershtein2006-bk}
found evidence for synergistic epistasis, in which the combined effect of
multiple deleterious mutations on individual fitness was larger than would be
expected from multiplying the independent observed effects of each individual
mutation. The scale of mutation scanning experiments continues to increase,
promising greater resolution of the fitness landscape in such model systems
that can be compared to evolutionary theory \citep{Otwinowski2018-zk}.

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theoretical
predictions that linked negatively selected mutations will display negative LD
due to interference \citep{Hill1966-gv} and that epistasis will drive expected
LD away from zero, multiple recent studies have used patterns of LD within
classes of putatively selected variants to infer modes of selective
interactions. \citet{Callahan2011-ac} observed that pairs of tightly linked
nonsynonymous mutations cluster together more than expected along lineages in
the \emph{Drosophilid} species complex, and that those clustered mutations tend
to preserve the charge of the protein and were in positive LD compared to pairs
of synonymous mutations at the same distance. From this, they proposed that
compensatory nonsynonymous variants regularly arise and are maintained. More
recently, \citet{Taverner2020-lk} replicated this finding across a diverse set
of genera, showing that such epistatic interactions are important in protein
evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. This was interpreted as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutation
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn}
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, suggesting interference and
dominance could be driving patterns of LD \citep{Garcia2021-zn} and
questioning whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

Many factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their low allele frequencies mean that measurements of LD and other diversity
statistics are very noisy. Second, comparisons are based on theory with
limiting assumptions, such as steady-state demography, simple selection and
interaction models, or unlinked loci. To generate predictions under more
complex models, we rely on expensive forward simulations. Such simulations can
help build intuition and be used to test inference methods, but they do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

\begin{figure}[tb!]
    %\internallinenumbers
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g., conditioned on observing
        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, decays
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts squared LD away from neutral expectations.
        (E) $\sigma_d^1$ (signed LD) is zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moment system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution of
two-locus haplotypes (\(\Psi_n\), Figure~\ref{fig:TwoLocusFS}) under a general
model of selective interactions. This sampling distribution stores the expected
density or observed counts of pairs of biallelic loci with each possible
haplotype configuration in a sample of size \(n\). Below, we compute
expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of mutations. It is not possible in this framework to
include the effects of additional linked selected mutations, such as background
selection, and individual-based forward simulations are still needed for such
scenarios (e.g., Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of
the distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution \(\Psi_n\). For pairs of biallelic loci, with
alleles labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
\(D=f_{AB}-f_A f_B\) is the standard covariance measure of LD, where \(f_{AB}\)
is the frequency of haplotypes carrying both \(A\) and \(B\), and \(f_A\) and
\(f_B\) are the marginal allele frequencies of \(A\) and \(B\). Here I focus on
low-order LD statistics and their decay with recombination distance, as these
are statistics that are commonly used to test for interactions between loci.
Instead of \(\E[D^2]\) and \(\E[D]\) I consider expectations for \(\sigma_d^2 =
\E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]\) and \(\sigma_d^1 =
\E[D]/\E[f_A(1-f_A)f_B(1-f_B)]\). Normalized statistics have two benefits: (1)
the mutation rate cancels so that expectations are robust to assumptions about
the per-base mutation rate, and (2) we can compare to analytic expectations for
these quantities under neutrality and constant population size
\citep{Ohta1971-yd}.

Below, I first consider the case of additive selection, Hill-Robertson
interference, and epistasis. I then explore the effect of dominance acting
within loci but without epistatis, and then describe a general diploid
selection model and consider gene-based dominance effects. I present results
for equal strengths of selection and dominance at each locus, but note that the
methods presented here allow for arbitrary and unequal selection and dominance
at the two loci. I also focus primarily on weak to moderate negative selection
(\(|2Ns| \approx 1 - 20\)), since this range of selection leads to the
strongest signals of interference (Figure~\ref{fig:HillRobertson}) and is the
parameter regime for which the numerical approach is most accurate.

@@

\begin{figure}[tb!]
    %\internallinenumbers
    \centering
    \includegraphics{../figures/hill_robertson}
    \caption{
        \textbf{Hill-Robertson interference.}
        Interference between pairs of selected mutations causes negative
        signed LD.
        (A) The expected normalized variance of $D$ (\(\sigma_d^2\))
        decreases below neutral expectations
        as the strength of negative selection increases.
        (B, C) For tightly linked loci (\(4Nr \lesssim 1\)), interference is most
        noticeable for pairs of mutations with \(s \approx 1/N\).
        At larger recombination distances (\(4Nr > 1\)), signed LD is most
        negative for somewhat stronger selection coefficients.
        Dashed lines show neutral expectations.
    }
    \label{fig:HillRobertson}
\end{figure}

\begin{figure}[tb!]
    %\internallinenumbers
    \centering
    \includegraphics{../figures/epistasis_prediction}
    \caption{
        \textbf{Additive selection and epistasis.}
        Left panels (A and C) show expectations for the decay of
        \(\sigma_d^2\) with recombination
        distance, and right panels (B and D) show expectations for the
        decay of \(\sigma_d^1\). Dashed lines
        show neutral expectations.
        For both weak (\(s=-1/2N\), A and B) and moderate
        (\(s=-10/2N\), C and D) selection,
        antagonistic epistasis (\(\epsilon < 0\))
        causes positive signed LD and increased $\sigma_d^2$ over
        a multiplicative model (\(\epsilon = 0\)),
        and synergistic epistasis (\(\epsilon > 0\)) results in negative signed
        LD beyond that of Hill-Robertson interference and decreased \(\sigma_d^2\).
    }
    \label{fig:epistasis}
\end{figure}

For mutations under additive selection (\(h=1/2\)) and no epistasis, we
recover the well known \citet{Hill1966-gv} interference result of negative LD
between selected mutations, which is strongest for pairs of mutations that have
selection coefficients \(\gamma = 2N_es = O(1)\), or \(s \approx 1/2N_e\) (Figure
\ref{fig:HillRobertson}).
For strongly deleterious mutations, LD is close to zero even
with tight linkage, as they almost always segregate at low enough frequencies
that they are unlikely to interfere with each other \citep{McVean2000-ox}.

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \ref{fig:epistasis}).
Synergistic epistasis (in which the effect of two mutations together is larger than
the product of each individual mutation's effect) results in negative LD while
antagonistic epistasis (in which the combined effect is less than the product of
independent effects) results in positive LD, and large nonzero LD can occur
even when epistasis is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations.
Even moderately deleterious mutations with
population-size-scaled selection coefficients of \(\gamma=-10\) show large mean
LD that extends to values of \(\rho\) much greater than 1
(Figure \ref{fig:epistasis}F; for humans, assuming
roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
More strongly deleterious interacting mutations are expected to show large
signed LD over much larger recombination distances.

@@

\begin{figure}[tb!]
    %\internallinenumbers
    \centering
    \includegraphics{../figures/dominance_roze}
    \caption{
        \textbf{The effect of dominance on LD.}
        (A, B) The strengths of selection and dominance interact in a nonlinear way
        to shape expected signed LD. For weakly to moderately selected mutations,
        as shown here, signed LD can be large and negative for tightly linked loci
        (e.g., \(\gamma=-5\), \(h=0\), and \(\rho < 1\)). However, this large signed
        LD decays with recombination distance faster under a model of recessivity
        than does signed LD under a model of additive selection and
        epistasis (Figure~\ref{fig:epistasis}).
        (C) Interference effects are most pronounced for recessive deleterious
        variants.
        (D, E) Recessive strongly deleterious mutations can have positive signed
        LD, as recently shown by \citet{Roze2021-cf}.
        However the dominance threshold at which \(\sigma_d^1\) switches from
        positive to negative depends on the strength of selection, and weakly
        selected mutations can show non-monotonic behavior as $h$ varies.
        Selection parameters of \(sh=-0.1\) imply extremely strong selection
        (\(h=0.05\) results in \(s=-0.2\) and \(\gamma=-400\) for homozygous
        diploids at a single locus). The numerical approach for \(\Psi_n\)
        cannot handle such strong selection.
    }
    \label{fig:dominance}
\end{figure}

The effect of non-additive selection on correlations between mutations has
received increased attention recently. For example, \citet{Garcia2021-zn} used
large-scale forward simulations to explore how dominance impacts patterns of
LD, showing that LD depends nonlinearly on the magnitudes of both selection
and dominance. \citet{Roze2021-cf} found an analytic expression
for LD between pairs of strongly deleterious mutations under steady-state
demography, showing that LD can be either positive or negative depending on the
strength of dominance.

The combined effect of the strengths of selection and dominance on interference
is indeed nontrivial (Figure~\ref{fig:dominance}A). Some parameter regimes can
cause large negative LD between negatively selected variants, with
moderately selected recessive variants showing stronger signals of interference
than additive selection (Figure~\ref{fig:dominance}B). Unlike epistatic
interactions, signed LD decays rapidly with increasing distance between loci
and is roughly zero for \(\rho \gg 1\). For weakly selected mutations
(\(|\gamma|\approx 1\)), there is no monotonic effect of the level of dominance
on negative LD, with both recessive and dominant pairs of mutations having more
negative LD than additive mutations.

Discrete simulations and the moment approach confirm the result from
\citet{Roze2021-cf}, that positive LD can occur for strong negative selection
and small values of \(h\) (Figure~\ref{fig:dominance}D, E). However, while the
analytic formula in \citet{Roze2021-cf} predicts that LD should be positive for
\(h<0.25\) and negative for \(h>0.25\), this appears to only hold in the limit
of strong selection (compare to Figure 1A in \citet{Roze2021-cf}). For moderate
to moderately strong selection, this threshold of \(h\) can be less than
\(0.25\), and LD is negative for all \(0\leq h \leq 1\) for weakly deleterious
mutations (Figure~\ref{fig:dominance}C).

\subsubsection{Arbitrary two-locus selection models}
\label{sec:general-selection}

\begin{figure}[tb!]
    %\internallinenumbers
    \centering
    \includegraphics{../figures/gene-based-compensatory}
    \caption{
        \textbf{Multiple modes of interactions can lead to large positive signed LD.}
        Both antagonistic epistasis (A and C, and which includes compensatory mutation
        models) and gene-based dominance (B and D) lead to large positive signed LD.
        Compensatory mutations (\(\epsilon \approx -1\)) also cause increased
        \(\sigma_d^2\) compared to neutral expectations (dashed black lines),
        while weaker antagonistic epistasis does not increase \(\sigma_d^2\) above
        neutral expectations (compare to Figure~\ref{fig:epistasis}C).
        Gene-based dominance instead causes lower \(\sigma_d^2\) than neutral
        expectations.
        While signed LD may be similar between different interaction models,
        other two-locus summaries of the data may help to distinguish between
        interaction models.
        }
    \label{fig:gene-based-compensatory}
\end{figure}

Beyond standard models of epistasis and dominance, a large family of selection
models can be specified by assigning unique fitness effects to each possible
diploid pair of haplotypes. Assuming the diploid genotype homozygous for the
ancestral alleles (\(ab/ab\)) has relative fitness 1, there are nine other
diploid two-locus genotypes that could be given unique fitnesses (Table
\ref{tab:selmodels}), noting that \(AB/ab\) and \(Ab/aB\) genotypes can have
differing selection coefficients.
